---
title: "C1"
author: "MartinezLeines"
date: "2025-04-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(writexl)
library(dplyr)
```

## Including Plots

You can also embed plots, for example:

```{r}
base_dir <- "~/R/LeinesMartinez_ENV797_TSA_FinalProject" # Update this
data_dir <- file.path(base_dir, "Data")
output_dir <- file.path(base_dir, "Output")

file1 <- "oil_production_EC_2007_2024.xlsx"
file2 <- "oil_price_2007-2024.xlsx"
file3 <- "annual_oil_production_1972-2023.xlsx"

file_path1 <- file.path(data_dir, file1)
file_path2 <- file.path(data_dir, file2)
file_path3 <- file.path(data_dir, file3)

oil_production <- read_excel(file_path1)
oil_prices <- read_excel(file_path2)
annual_oil_production <- read_excel(file_path3)

```


```{r}
oil_data <- oil_production |>
  left_join(oil_prices, by = "date") |>
  select(-month.y, -year.y, -month.x, -year.x) |>
  mutate(date = as.Date(date)) |>
  select(date, everything())


output_file1 <- file.path(output_dir, "oil_data_2007_2024.xlsx")
write_xlsx(oil_data, path = output_file1)
  

```


```{r}
output_file3 <- file.path(output_dir, "oil_data_72_2023.xlsx")
annual_data_72_2023 <- annual_oil_production |>
  slice(-1:-14) |>               
  select(2, 8) |>                     
  setNames(c("year", "annual_production")) |> 
  filter(!(is.na(year) & is.na(annual_production))) %>%
  mutate(
    year = as.numeric(year),
    annual_production = as.numeric(annual_production),
    annual_production = as.numeric(annual_production) * 1000 
  )

write_xlsx(annual_data_72_2023, path = output_file3)

```

```{r}
#I forced to 0 all the NA in our processed data
oil_data1<-oil_data %>%
  replace_na(list(barrels_b043 = 0, b043_share = 0))
```

```{r}
#Trim the table to include only columns you need
noil <- ncol(oil_data1)-2
nobs <- nrow(oil_data1) 
```


```{r}
p1 <-ggplot(oil_data, aes(x = date, y = barrels_total)) +
  geom_line() +
  labs(title = "Monthly data over time", x = "Date", y = "Oil production") +
  theme_classic()

print(p1)

p2 <-ggplot(oil_data1, aes(x = date, y = barrels_other)) +
  geom_line() +
  labs(title = "Monthly data over time", x = "Date", y = "Oil production") +
  theme_classic()

print(p2)

p3 <-ggplot(oil_data1, aes(x = date, y = barrels_b043)) +
  geom_line() +
  labs(title = "Monthly data over time", x = "Date", y = "Oil production") +
  theme_classic()

print(p3)
```



```{r message=FALSE, warning=FALSE}
#I corrected the columns
ts_oil_other <- ts(oil_data1[,4],
                           start=c(2013,1),
                           frequency=12) 

ts_oil_b043 <- ts(oil_data1[,3],
                           start=c(2013,1),
                           frequency=12) 

ts_oil_total <- ts(oil_data1[,2],
                           start=c(2013,1),
                           frequency=12) 

```


```{r}
p4 <- autoplot(ts_oil_other)
print(p4)

p5 <- autoplot(ts_oil_b043)
print(p5)

p6 <- autoplot(ts_oil_total)
print(p6)
```

```{r}
#Plot
autoplot(ts_oil_total, color="darkblue")+
  autolayer(ts_oil_b043,series="043",color="green")+
  autolayer(ts_oil_other,series="Other",color="red")
```

```{r}
annual_ts <- ts(annual_data_72_2023[,2],
                start = c(1972, 1),
                frequency = 1)
p7 <- autoplot(annual_ts)
print(p7)
```

```{r}
#training
annual_ts_train <- window(annual_ts,
                          start= c(2000,1),
                end= c(2020, 1),
                frequency = 1)

ts_daily_test <- window(annual_ts,
                start= c(2021, 1),
                frequency = 1)
```

#Model 1: ARIMA
```{r ARIMA}
#Model 1: ARIMA
# Fit an ARIMA model to the annual time series and forecast for 3 years
model_arima <- auto.arima(annual_ts_train)
forecast_arima <- forecast(model_arima, h = 3)
print(forecast_arima)

# Plot the forecast
plot(forecast_arima)


#Plot model + observed data
autoplot(annual_ts) +
  autolayer(forecast_arima, series="AutoArima",PI=FALSE) +
  ylab("Annual oil production Ecuador")

```

### Testing Model 2:  MEAN_seas
```{r MEAN, echo=TRUE, message=FALSE, warning=FALSE}
#Model 2: Arithmetic mean on original data
MEAN_seas <- meanf(y = annual_ts_train, h = 3)
plot(MEAN_seas)

autoplot(annual_ts) +
  autolayer(MEAN_seas, series="MEAN",PI=FALSE) +
  ylab("Production")

```

```{r}
# Model 3: ETS (Exponential Smoothing without seasonality)
model_ets <- ets(annual_ts_train)
forecast_ets <- forecast(model_ets, h = 3)

autoplot(annual_ts) +
  autolayer(forecast_ets, series="ETS",PI=FALSE) +
  ylab("Production")
```

```{r}
# Model 4: Holt's Linear Trend method
model_holt <- holt(annual_ts_train, h = 3)
forecast_holt <- forecast(model_holt, h = 3)

autoplot(annual_ts) +
  autolayer(forecast_holt, series="HOLT",PI=FALSE) +
  ylab("Production")
```
```{r}
#Model 1: ARIMA
ARIMA_scores <- accuracy(forecast_arima$mean,ts_daily_test)  #store the performance metrics

#Model 2: Arithmetic mean 
MEAN_scores <- accuracy(MEAN_seas$mean,ts_daily_test)

# Model 3:  ETS
ETS_scores <- accuracy(forecast_ets$mean,ts_daily_test)

# Model 4:  HOLT
HOLT_scores <- accuracy(forecast_holt$mean,ts_daily_test)

```

### Compare performance metrics

Now we will create a data frame that combines performance metrics for all the three models. You can choose one metric to help you choose among models. For example let's say we want the model with lowest RMSE.

```{r}
#create data frame
models_scores <- as.data.frame(rbind(MEAN_scores, ARIMA_scores,ETS_scores,HOLT_scores ))
row.names(models_scores) <- c("MEAN", "ARIMA","ETS","HOLT")

#choose model with lowest RMSE
best_model_index <- which.min(models_scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(models_scores[best_model_index,]))                       
                            
```


SARIMA was the best fit for the seasonal data. If you want generate a table to compare model accuracy and help visualize the results here is a suggestion on how to include a table on your Rmd report. You can use the `kable_styling(latex_options="striped")` to highlight the model that leads to minimum RMSE.

```{r}
kbl(models_scores, 
      caption = "Forecast Accuracy for Annual Data",
      digits = array(5,ncol(models_scores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(models_scores[,"RMSE"]))
```


```{r}
# Create a training set using data from 2020 to 2023
annual_ts_2023 <- window(annual_ts, start = c(2000, 1), end = c(2023, 1))

# Fit an ARIMA model on the updated training data
model_arima_2023 <- auto.arima(annual_ts_2023)

# Forecast the next 4 years (2024–2027)
forecast_2024_2027 <- forecast(model_arima_2023, h = 4)

# Print and plot the forecast
print(forecast_2024_2027)
plot(forecast_2024_2027)

```

```{r}
# 1. Filter existing data from 2000 to 2023
expected_production <- annual_data_72_2023 %>%
  filter(year >= 2000, year <= 2023)

# 2. Create a data frame for 2024–2027: data from https://www.primicias.ec/economia/plan-hidrocarburifero-inversiones-petroleo-gobierno-noboa-88061/
daily_values <- c(475.27, 508.09, 600.72, 539.252)  # daily production according to Ecuador's gov
future_years <- 2024:2027

df_future <- data.frame(
  year = future_years,
  annual_production = daily_values * 1000 * 365
)

# 3. Combine the filtered data with the new rows
expected_production <- bind_rows(expected_production, df_future)

annual_exp_ts <- ts(expected_production[,2],
                start = c(2000, 1),
                frequency = 1)


# 6. Plot the ARIMA forecast and government expected production together
autoplot(forecast_2024_2027) +
  autolayer(annual_exp_ts, series = "Government Expected", linetype = "dashed", color = "darkblue") +
  xlab("Year") +
  ylab("Annual Production") +
  ggtitle("ARIMA Forecast vs. Government Expected Production (2024-2027)") +
  guides(colour = guide_legend(title = "Series"))

```