---
title: "C1"
author: "MartinezLeines"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(writexl)
library(dplyr)
```

## Including Plots

You can also embed plots, for example:

```{r}
base_dir <- "D:/Geani/Box/Home Folder gnl13/Private/1 Academics/3 Time series/LeinesMartinez_ENV797_TSA_FinalProject" # Update this
data_dir <- file.path(base_dir, "Data")
output_dir <- file.path(base_dir, "Output")

file1 <- "oil_production_EC_2007_2024.xlsx"
file2 <- "oil_price_2007-2024.xlsx"
file3 <- "annual_oil_production_1972-2023.xlsx"
file4 <- "oil_data_2007_2024.xlsx"

file_path1 <- file.path(data_dir, file1)
file_path2 <- file.path(data_dir, file2)
file_path3 <- file.path(data_dir, file3)
file_path4 <- file.path(data_dir, file4)

oil_production <- read_excel(file_path1)
oil_prices <- read_excel(file_path2)
annual_oil_production <- read_excel(file_path3)
oil_data <- read_excel(file_path4)
```
```{r}
#I forced to 0 all the NA in our processed data
oil_data1<-oil_data |>
  mutate(date=as.Date(date))|>
  replace_na(list(barrels_b043 = 0, b043_share = 0))
```

```{r}
#Trim the table to include only columns you need
noil <- ncol(oil_data1)-2
nobs <- nrow(oil_data1) 
```


```{r}
p1 <-ggplot(oil_data1, aes(x = date, y = barrels_total)) +
  geom_line(color = "blue") +
  labs(title = "Monthly Oil Production in Ecuador", x = "Date", y = "Oil production") +
  theme_classic()

print(p1)

p2 <-ggplot(oil_data1, aes(x = date, y = wti_oil_price)) +
  geom_line(color = "darkgreen") +
  labs(title = "Monthly WTI Prices", x = "Date", y = "WTI Price")+
  theme_classic()

print(p2)

p3<- ggplot(oil_data1, aes(x = date, y = barrels_b043)) +
  geom_line(color = "red") +
  labs(title = "Monthly Block 43 ITT Production", x = "Date", y = "Block 43 ITT Production") +
  theme_classic()

print(p3)



```



```{r message=FALSE, warning=FALSE}
#I corrected the columns
ts_oil_total <- ts(oil_data1[,2],
                           start=c(2007,1),
                           frequency=12) 

ts_oil_b043 <- ts(oil_data1[,3],
                           start=c(2007,1),
                           frequency=12) 

ts_oil_price <- ts(oil_data1[,6],
                           start=c(2007,1),
                           frequency=12) 

```


```{r}
p4 <- autoplot(ts_oil_total)
print(p4)

p5 <- autoplot(ts_oil_b043)
print(p5)

p6 <- autoplot(ts_oil_price)
print(p6)
```

```{r}
# ACF and PACF for Production
Acf(ts_oil_total, lag.max = 40, main = "ACF of Production")
Pacf(ts_oil_total, lag.max = 40, main = "PACF of Production")
```



```{r}
# Decompose the Production series (additive decomposition)
decomp_prod <- decompose(ts_oil_total, type = "additive")
plot(decomp_prod)

# Remove the seasonal component
deseasonal_prod <- seasadj(decomp_prod)

# Compare the original and deseasonalized series
autoplot(ts_oil_total, series = "Original") +
  autolayer(deseasonal_prod, series = "Deseasonalized") +
  labs(title = "Production: Original vs. Deseasonalized")
```



```{r}
# Perform ADF test on the deseasonalized Production series
adf_result <- adf.test(deseasonal_prod, alternative = "stationary")
print(adf_result)
```
```{r}
# Difference the deseasonalized series
diff_prod <- diff(deseasonal_prod, differences = 1)

# Plot the differenced series
autoplot(diff_prod, main = "Differenced Deseasonalized Production") +
  labs(x = "Time", y = "Differenced Production")
```
```{r}
# Check stationarity again
adf_result_diff <- adf.test(diff_prod, alternative = "stationary")
print(adf_result_diff)
```
```{r}
# ACF and PACF for the differenced series
Acf(diff_prod, lag.max = 40, main = "ACF of Differenced Series")
Pacf(diff_prod, lag.max = 40, main = "PACF of Differenced Series")
```

> Analysis ACF: The spike at lag 1 appears somewhat negative (below the lower blue line), suggesting some immediate negative correlation with the first lag. That sometimes indicates an MA(1) effect or a negative AR(1) coefficient.

>Analysis PACF: There’s a notable negative spike at lag 1, and possibly some mild spikes around lag 3 or 4.

```{r}
# to the deseasonalized Production series
model_auto <- auto.arima(deseasonal_prod)
summary(model_auto)
```
```{r}
checkresiduals(model_auto)
```
> Analysis: the residuals mostly fluctuate around zero. It has outliers due to COVID and other event around 2022.It could be good to add them as dummies.

>Analysis ACF:  ACF looks mostly within the bounds, indicating little to no autocorrelation in the residuals. That’s a sign your ARIMA(0,1,2) model is adequately capturing time‐dependent structure. There are 3 main spikes at specific lags. 

>Analysis Residual histogram: The histogram looks centered near zero even when there are some outliers.

> Analysis test: p‐value (< 0.05) in the Ljung‐Box test indicates that there is still significant autocorrelation in residuals. The ARIMA(0,1,2) model hasn’t fully captured the time‐dependent structure of the data.

```{r}
Acf(residuals(model_auto))
Pacf(residuals(model_auto))

```
```{r}
model_sarima <- auto.arima(
  ts_oil_total,
  seasonal = TRUE
)

summary(model_sarima)
```

```{r}
checkresiduals(model_sarima)
```
> Analysis test: the very low Ljung–Box p-value indicates that its residuals are not white noise. The  SARIMA(2,1,1)(0,0,1)[12] model on the total series isn’t fully capturing the dynamics.


```{r}
Acf(residuals(model_sarima))
Pacf(residuals(model_sarima))
```


```{r}
# For example, create a matrix of external regressors from your oil_data
# Make sure the regressor vectors are of the same length as your Production series
xreg_all <- cbind(WTI = oil_data1$wti_oil_price, Block43 = oil_data1$barrels_b043)
xreg_price <- oil_data1$wti_oil_price
```

```{r}
# Fit the SARIMAX model using auto.arima with xreg
model_sarimax <- auto.arima(deseasonal_prod, xreg = xreg_all, seasonal = TRUE)
summary(model_sarimax)


```

```{r}
# Check residuals for the SARIMAX model
checkresiduals(model_sarimax)

```
> Analysis: p‐value = 0.1569 (> 0.05) suggests no significant autocorrelation remains in the residuals.The SARIMAX model with regressors yielded residuals that appear closer to white noise.


```{r}
model_sarimax_orig <- auto.arima(
  ts_oil_total,
  xreg = cbind(oil_data1$wti_oil_price, oil_data1$barrels_b043),
  seasonal = TRUE, 
  stepwise = FALSE,       # often better for thorough search
  approximation = FALSE
)
summary(model_sarimax_orig)

```

```{r}
checkresiduals(model_sarimax_orig)
```
```{r}

```




#Train/Test

```{r}
# Restrict the production series to data until the end of 2023
ts_oil_total_2023 <- window(ts_oil_total, end = c(2023, 12))
deseasonal_prod_2023 <- window(deseasonal_prod, end = c(2023, 12))
oil_data_2023 <- oil_data1[oil_data1$date <= as.Date("2023-12-31"), ]
xreg_all_2023 <- cbind(WTI = oil_data_2023$wti_oil_price, Block43 = oil_data_2023$barrels_b043)
price_regressor_2023 <- oil_data_2023$wti_oil_price

# Define forecast horizon
n_for <- 12
n_total <- length(ts_oil_total_2023)
n_total_des <- length(deseasonal_prod_2023)

# For Model A (total series)
ts_train_A <- window(ts_oil_total_2023, end = time(ts_oil_total_2023)[n_total - n_for])
ts_test_A <- window(ts_oil_total_2023, start = time(ts_oil_total_2023)[n_total - n_for + 1])

# For Model B (deseasonalized series)

ts_train_B <- window(deseasonal_prod_2023, end = time(deseasonal_prod_2023)[n_total_des - n_for])
ts_test_B <- window(deseasonal_prod_2023, start = time(deseasonal_prod_2023)[n_total_des - n_for + 1])

ts_train_C <- window(ts_oil_total, end = time(ts_oil_total)[n_total - n_for])
ts_test_C <- window(ts_oil_total, start = time(ts_oil_total)[n_total - n_for + 1])


# For exogenous regressors:
xreg_train <- xreg_all_2023[1:(n_total - n_for), ]
xreg_test <- xreg_all_2023[(n_total - n_for + 1):n_total, ]
price_train <- price_regressor_2023[1:(n_total - n_for)]
price_test <- price_regressor_2023[(n_total - n_for + 1):n_total]

# Forecast horizon
h <- n_for
```

#Model 1

```{r}
# Model A: Baseline SARIMA on total production
model_1_train <- auto.arima(ts_train_A, seasonal = TRUE)

# Forecast for Model A
forecast_1 <- forecast(model_1_train, h = h)

print(forecast_1)

# Plot the forecast
plot(forecast_1)


#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_1, series="Sarima",PI=FALSE) +
  ylab("Monthly oil production Ecuador")
```

#Model 2:

```{r}

# Model B: SARIMAX on deseasonalized production with regressors
model_2_train <- auto.arima(ts_train_B, xreg = xreg_train, seasonal = TRUE)

# Forecast for Model B (with xreg)
forecast_2 <- forecast(model_2_train, xreg = xreg_test, h = h)

print(forecast_2)

# Plot the forecast
plot(forecast_2)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_2, series="Sarimax",PI=FALSE) +
  ylab("Monthly oil production Ecuador")
```

#Model 3

```{r}
# Model 3: ARIMAX on deseasonalized production with regressors
model_3_train <- auto.arima(ts_train_A, xreg = xreg_train, seasonal = TRUE)

# Forecast for Model C (with xreg)
forecast_3 <- forecast(model_3_train, xreg = xreg_test, h = h)

print(forecast_3)

# Plot the forecast
plot(forecast_3)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_3, series="Arimax",PI=FALSE) +
  ylab("Monthly oil production Ecuador")
```

#Model 4

```{r}
# Model 4
model_4_train <- meanf(ts_train_A, h = h)

# Plot the forecast
plot(model_4_train)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(model_4_train, series="Mean",PI=FALSE) +
  ylab("Monthly oil production Ecuador")
```
```{r}
# Model 5: 
model_5_train <- ets(ts_train_A, model = "ANN")

# Forecast for Model 5
forecast_5 <- forecast(model_5_train, h = h)

print(forecast_5)

# Plot the forecast
plot(forecast_5)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_5, series="ETS",PI=FALSE) +
  ylab("Monthly oil production Ecuador")
```


```{r}
# Model 6: 
model_6_train <- tbats(ts_train_A)

# Forecast for Model 6
forecast_6 <- forecast(model_6_train, h = h)

print(forecast_6)

# Plot the forecast
plot(forecast_6)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_6, series="TBATS",PI=FALSE) +
  ylab("Monthly oil production Ecuador")



```
```{r}
# Model 7: 
model_7_train <- nnetar(ts_train_A)

# Forecast for Model 7
forecast_7 <- forecast(model_7_train, h = h)

print(forecast_7)

# Plot the forecast
plot(forecast_7)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_7, series="Neural Network",PI=FALSE) +
  ylab("Monthly oil production Ecuador")

```

```{r}
# Model 8: 
model_8_train <- StructTS(ts_train_A, type = "BSM")

# Forecast for Model 8
forecast_8 <- forecast(model_8_train, h = h)

print(forecast_8)

# Plot the forecast
plot(forecast_8)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_8, series="StructTS",PI=FALSE) +
  ylab("Monthly oil production Ecuador")

```


#Model 9

```{r}
# Model 8: 
model_9_train <- auto.arima(ts_train_A, 
                             xreg = price_train, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)

# Forecast for Model 8
forecast_9 <- forecast(model_9_train, h = h, xreg = price_test)

print(forecast_9)

# Plot the forecast
plot(forecast_9)

#Plot model + observed data
autoplot(ts_oil_total) +
  autolayer(forecast_9, series="ARIMAX (Price Only)",PI=FALSE) +
  ylab("Monthly oil production Ecuador")

```



```{r}

#Model 1
SARIMA_scores <- accuracy(forecast_1$mean, ts_test_A)


#Model 2
SARIMAX_scores <- accuracy(forecast_2$mean, ts_test_B)

#Model 3
ARIMAX_scores <- accuracy(forecast_3$mean, ts_test_A)

#Model 4
Mean_scores <- accuracy(model_4_train$mean, ts_test_A)

#Model 5
ETS_scores <- accuracy(forecast_5$mean, ts_test_A)

#Model 6
TBATS_scores <- accuracy(forecast_6$mean, ts_test_A)

#Model 7
NN_scores <- accuracy(forecast_7$mean, ts_test_A)

#Model 8
StructTS_scores <- accuracy(forecast_8$mean, ts_test_A)


#Model 9
Arimax_p_scores <- accuracy(forecast_9$mean, ts_test_A)
```


```{r}
# Combine in a table for easy comparison
models_scores <- as.data.frame(rbind(SARIMA_scores, SARIMAX_scores,ARIMAX_scores,
                                     Mean_scores,ETS_scores, TBATS_scores,
                                     NN_scores,StructTS_scores, Arimax_p_scores ))
row.names(models_scores) <- c("SARIMA", "SARIMAX", "ARIMAX",
                              "Mean","ETS", "TBATS",
                                     "NN","StructTS", "Arimax_p")
print(models_scores)

#choose model with lowest RMSE
best_model_index <- which.min(models_scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(models_scores[best_model_index,])) 
```



```{r}

autoplot(ts_oil_total) +
  autolayer(forecast_1, series = "Model 1: SARIMA", PI = FALSE) +
  autolayer(forecast_2, series = "Model 2: SARIMAX", PI = FALSE) +
  autolayer(forecast_3, series = "Model 3: ARIMAX", PI = FALSE) +
  autolayer(model_4_train, series = "Model 4: Mean", PI = FALSE) +
  autolayer(forecast_5, series = "Model 5: ETS", PI = FALSE) +
  autolayer(forecast_6, series = "Model 6: TBATS", PI = FALSE) +
  autolayer(forecast_7, series = "Model 7: NN", PI = FALSE) +
  autolayer(forecast_8, series = "Model 8: StructTS", PI = FALSE) +
  autolayer(forecast_8, series = "Model 9: Arimax_p", PI = FALSE) +
  ggtitle("Forecast Comparison") +
  xlab("Time") + ylab("Production")+
  guides(colour=guide_legend(title="Forecast"))

```
```{r}
kbl(models_scores, 
      caption = "Forecast Accuracy for Monthly Data",
      digits = array(9,ncol(models_scores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(models_scores[,"RMSE"]))
```

#Scenario Analysis

```{r}
#TBATS—the best‐performing model among your nine—to conduct the scenario analysis. Because TBATS is a univariate model (it does not accept exogenous regressors), we must estimate the Block 43 contribution separately. One common approach is to compute the average monthly production from Block 43 during a recent period and then “remove” that contribution from the TBATS baseline forecast to simulate a shutdown. In other words, your shutdown scenario forecast equals the TBATS baseline forecast minus the estimated Block 43 production (or a ramp‐down profile if desired).

#The code proceeds as follows:

#Fit a TBATS model to the total production series through 2023.
#Generate a baseline forecast using TBATS.
#Compute the average Block 43 production over the last 12 months (or use a ramp‐down vector).
#Create a “shutdown scenario” forecast by subtracting that average from the TBATS forecast.
#Compute and plot the production gap.


```


```{r}
#Fit TBATS Model on Total Production ----
tbats_model <- tbats(ts_oil_total_2023)
forecast_baseline <- forecast(tbats_model, h = h)
print(forecast_baseline)

```


```{r}
# ---- Step 4: Estimate Block 43 Contribution ----
# Here, we compute the average monthly production from Block43 over the last 12 months.
average_block43 <- mean(tail(oil_data_2023$barrels_b043, 12))
cat("Average monthly Block 43 production:", average_block43, "\n")

# Alternatively, you could create a ramp-down vector if you expect a gradual shutdown.
# For a simple case, we use a constant value:
block43_shutdown <- rep(average_block43, h)  # this will be subtracted from the baseline

```


```{r}


# ---- Step 5: Create Shutdown Scenario Forecast ----
# The shutdown scenario forecast is computed by subtracting Block43's contribution.
forecast_shutdown <- forecast_baseline
forecast_shutdown$mean <- forecast_baseline$mean - block43_shutdown

# ---- Step 6: Compute Production Gap ----
production_gap <- forecast_baseline$mean - forecast_shutdown$mean
cat("Production gap (per month):\n")
print(production_gap)

```


```{r}


# ---- Step 7: Plot the Forecast Scenarios ----
autoplot(forecast_baseline) +
  autolayer(forecast_shutdown, series = "Shutdown Scenario", PI = FALSE) +
  ggtitle("TBATS Forecast: Baseline vs. Block 43 Shutdown Scenario") +
  xlab("Time") + ylab("Total Oil Production") +
  guides(colour = guide_legend(title = "Scenario"))


```

```{r}
# ---- Step 8: Plot the Production Gap ----


gap_df <- data.frame(
  Time = time(production_gap),
  Gap  = as.numeric(production_gap)
)

# Convert numeric time to a "yearmon" format, then to a Date
gap_df$TimeDate <- as.Date(as.yearmon(gap_df$Time))

ggplot(gap_df, aes(x = TimeDate, y = Gap)) +
  geom_line(color = "red", size = 1) +
  ggtitle("Estimated Production Gap due to Block 43 Shutdown") +
  xlab("Date") +
  ylab("Production Gap") +
  scale_x_date(
    date_breaks = "3 month",    # or "3 months", "1 year", etc.
    date_labels = "%b-%Y"       # e.g., "Jan-2024"
  ) +
  theme_minimal()
```



```{r}
# Instead of using a constant shutdown vector:
# block43_shutdown <- rep(average_block43, h)

# Create a ramp-down vector: 
# For example, if you want production to decline linearly from average_block43 to 0:
block43_shutdown <- seq(from = average_block43, to = 0, length.out = h)

# Now create the shutdown scenario forecast by subtracting the ramp-down vector:
forecast_shutdown_ramp <- forecast_baseline
forecast_shutdown_ramp$mean <- forecast_baseline$mean - block43_shutdown

# Compute the production gap for the ramp-down scenario
production_gap_ramp <- forecast_baseline$mean - forecast_shutdown_ramp$mean


gap_df_ramp <- data.frame(
  Time = time(production_gap_ramp),
  Gap  = as.numeric(production_gap_ramp)
)

# Convert numeric time to a "yearmon" format, then to a Date
gap_df_ramp$TimeDate <- as.Date(as.yearmon(gap_df_ramp$Time))


# Plot the production gap for the ramp-down scenario:
ggplot(gap_df_ramp, aes(x = TimeDate, y = Gap)) +
  geom_line(color = "red", size = 1) +
  ggtitle("Estimated Production Gap with Ramp-Down of Block 43") +
  xlab("Time") + ylab("Production Gap")
```
```{r}

forecast_obj <- forecast(tbats_model, h = 24)  # forecast 2 years ahead, for example
# Aggregate the monthly forecast to annual totals.
# 'nfrequency = 1' converts the series to annual frequency.
annual_forecast <- aggregate(forecast_obj$mean, nfrequency = 1, FUN = sum)

print(annual_forecast)
```

```{r}
autoplot(annual_forecast, series = "Aggregated Forecast") +
  autolayer(annual_ts, series = "Historical Annual Data", PI = FALSE) +
  ggtitle("Annual Aggregated Forecast vs. Historical Annual Data") +
  xlab("Year") + ylab("Annual Total Oil Production") +
  guides(colour = guide_legend(title = "Series"))
```

